---
title: "Mitigating Class Imbalance with Tidymodels: Attrition Example"
format:
  html:
    self-contained: true
    toc: true
    toc-location: left
    toc-depth: 3
    code-fold: false
    code-tools: true
    theme: minty
    number-sections: false
execute:
  warning: false
  message: false
---


## Libraries

```{r}
library(tidymodels)
library(themis)
library(probably)
library(hardhat)
library(dplyr)
library(ggplot2)
library(modeldata)
```

## Load and Explore Data

```{r}
data(attrition)
attrition <- attrition %>%
  mutate(Attrition = factor(Attrition, levels = c("No", "Yes")))

```

- check what the data is about

- check the class imbalance

- check your predictors


## Deal with class imbalance

- pre-processing methods for class imbalance (e.g., upsampling, downsampling, SMOTE)

- algorithms that can handle class imbalance (cost-sensitive learning). Define weights for observations given the class they are in:

  - `importance_weights()`: used to create a new column in the data frame that contains the weights for each observation.
  - `step_add_weights()`: used to add weights to the recipe.


- threshold tuning: selecting the optimal threshold for classification


## Learning flow

- split the data

- create the multiple recipes you need

- create the model specifications for what you are willing to use

- create the workflows

- fit and assess the performance

- focus on F1, `precision` and `recall` (sensitivity) metrics

## Tasks

- What is the best setup? 

- How things improve over the solutions you get when imbalance is not dealt with?

- bonus: what happens if the imbalance is more severe? (e.g., 95% vs 5%)


