---
title: "Customer Churn Analysis and Prediction"
author: "Mohammad_Ehsani"
format: 
  html:
    theme: minty
    code-fold: true
    code-tools: true
    embed-resources: true
    page-layout: full
    toc: true
    toc-title: "Tasks"
    toc-location: left
    toc-expand: 2
knitr:
  opts_chunk:
    collapse: true
    message: false
    warning: false
    code-fold: false
---

```{r setup, include=FALSE}
# Load required libraries
library(tidyverse)
library(tidymodels)
library(skimr)
library(janitor)
library(vip)
library(iml)
library(pdp)
library(lime)
library(workflowsets)
library(knitr)
library(ggplot2)
library(dplyr)
library(RColorBrewer)
library(patchwork)
library(gridExtra)
library(ranger)
library(caret)
library(kernlab)
library(pROC)
```

# Project Objective

Analyze and predict customer churn using a structured machine learning workflow. Compare multiple models (linear, ensemble, and SVM) for both performance and interpretability.

# Load Dataset

```{r}
# Read the dataset
df <- read.csv("dataset/Databel - Data.csv")
```

# Dataset Dimensions

```{r}
# Display number of rows and columns
dim(df)
```

# Dataset Overview

```{r}
# Summary statistics
skim(df)
```

```{r}
# Structure and data types
str(df)
```

# Data Preprocessing

```{r}
# Clean column names (convert to lowercase and use underscores)
df <- clean_names(df)
```

# Missing Value Analysis

```{r}
# Custom function to detect missing values (including "NA" as string, empty string, or space)
missing_check <- function(x) {
  sum(is.na(x) | x == "" | x == " " | x == "NA")
}

# Build summary table of missing values
missing_summary <- data.frame(
  Column = names(df),
  Custom_Missing_Count = sapply(df, missing_check),
  Total_Records = nrow(df),
  Missing_Percent = round(sapply(df, missing_check) / nrow(df) * 100, 2)
)

# Filter only columns with missing values
missing_summary <- missing_summary[missing_summary$Custom_Missing_Count > 0, ]

# Display the missing value summary
print(missing_summary)
```

# Visualizing Missing Values

```{r Pie_Chart_for_missing_values, fig.width=12, fig.height=6}
# Function to prepare pie chart data for missing vs. present values
prepare_pie_data <- function(column_name) {
  row <- missing_summary %>% filter(Column == column_name)
  data.frame(
    Status = c("Missing", "Present"),
    Percent = c(row$Missing_Percent, 100 - row$Missing_Percent)
  ) %>%
    mutate(Label = paste0(Status, ": ", round(Percent, 1), "%"))
}

# Prepare pie chart data for 'churn_category' and 'churn_reason'
data_category <- prepare_pie_data("churn_category")
data_reason <- prepare_pie_data("churn_reason")

# Pie chart for 'churn_category'
p1 <- ggplot(data_category, aes(x = "", y = Percent, fill = Status)) +
  geom_col(width = 1) +
  coord_polar(theta = "y") +
  geom_text(aes(label = Label), position = position_stack(vjust = 0.5)) +
  scale_fill_manual(values = c("Missing" = "#D55E00", "Present" = "#56B4E9")) +
  labs(title = "Churn Category: Missing vs Present") +
  theme_void() + theme(legend.position = "none")

# Pie chart for 'churn_reason'
p2 <- ggplot(data_reason, aes(x = "", y = Percent, fill = Status)) +
  geom_col(width = 1) +
  coord_polar(theta = "y") +
  geom_text(aes(label = Label), position = position_stack(vjust = 0.5)) +
  scale_fill_manual(values = c("Missing" = "#D55E00", "Present" = "#56B4E9")) +
  labs(title = "Churn Reason: Missing vs Present") +
  theme_void() + theme(legend.position = "none")

# Arrange charts side by side
grid.arrange(p1, p2, ncol = 2)
```

# Categorical Variable Summary

```{r}
# Select character or factor columns
cat_df <- df %>% select(where(~ is.character(.x) || is.factor(.x)))

# Build summary table for categorical features
cat_summary <- cat_df %>%
  summarise(across(everything(), list(
    count = ~ sum(!is.na(.)),
    unique = ~ n_distinct(.),
    top = ~ names(sort(table(.), decreasing = TRUE)[1]),
    freq = ~ max(table(.))
  ), .names = "{.col}_{.fn}"))

# Reshape into readable format
summary_table <- tibble(
  Column = names(cat_df),
  count = as.numeric(cat_summary[1, seq(1, ncol(cat_summary), 4)]),
  unique = as.numeric(cat_summary[1, seq(2, ncol(cat_summary), 4)]),
  top = as.character(cat_summary[1, seq(3, ncol(cat_summary), 4)]),
  freq = as.numeric(cat_summary[1, seq(4, ncol(cat_summary), 4)])
)

# View summary
print(summary_table)
```

# Churn Summary

```{r}
# Total number of unique customers
total_customers <- n_distinct(df$customer_id)

# Total churned customers (assuming "Yes" = churned)
total_churned <- sum(df$churn_label == "Yes", na.rm = TRUE)

# Churn rate
churn_rate <- round((total_churned / total_customers) * 100, 2)

# Output
cat("Total Number of Customers:", total_customers, "\n")
cat("Total Number of Churned Customers:", total_churned, "\n")
cat("Rate of Churned Customers:", churn_rate, "%\n")
```

# Top Churn Reasons

```{r}
# Filter only churned customers
churned <- df %>%
  filter(churn_label == "Yes") %>%
  select(churn_reason, churn_category)

# Count churn reasons and sort
churn_reason_summary <- churned %>%
  count(churn_reason, name = "count") %>%
  arrange(desc(count))

# Show top 5 churn reasons
head(churn_reason_summary, 5)
```

# Churn Reason Distribution (Bar Plot)

```{r Count_of_All_Churn_Reasons, fig.width=12, fig.height=6}
# Replace missing/blank churn reasons with "Unknown"
df$churn_reason <- ifelse(is.na(df$churn_reason) | trimws(df$churn_reason) == "", "Unknown", df$churn_reason)

# Count reasons for churned customers
churn_reason_filtered <- df %>%
  filter(churn_label == "Yes") %>%
  count(churn_reason, name = "count") %>%
  arrange(desc(count))

# Plot bar chart
ggplot(churn_reason_filtered, aes(x = count, y = reorder(churn_reason, count), fill = count)) +
  geom_bar(stat = "identity") +
  scale_fill_viridis_c(option = "plasma", direction = -1) +
  labs(
    title = "Count of All Churn Reasons",
    x = "Count",
    y = "Churn Reason"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    axis.title = element_text(face = "bold", size = 13),
    legend.position = "none"
  )
```

# Churn vs. Retention by Account Length

```{r}
# Churned customers grouped by account length
churned <- df %>%
  filter(churn_label == "Yes") %>%
  group_by(account_length_in_months) %>%
  summarise(Churned = n(), .groups = "drop")

# Retained customers grouped by account length
retained <- df %>%
  filter(churn_label == "No") %>%
  group_by(account_length_in_months) %>%
  summarise(Not_Churned = n(), .groups = "drop")

# Merge churned and retained summaries
merged_df <- inner_join(churned, retained, by = "account_length_in_months")

# Display table
kable(
  head(merged_df, 10),
  caption = "Churned vs Not Churned by Account Length (in months)",
  col.names = c("Account Length (in months)", "Churned", "Not Churned")
)
```

# Line Plot: Churn by Account Length

```{r exact-lineplot-account-length, fig.width=12, fig.height=6}
# Rename column for clarity
colnames(merged_df)[1] <- "Account.Length"

# Convert to long format
merged_long <- merged_df %>%
  pivot_longer(cols = c("Churned", "Not_Churned"),
               names_to = "Status", values_to = "Count")

# Line plot
ggplot(merged_long, aes(x = Account.Length, y = Count, color = Status, linetype = Status)) +
  geom_line(size = 1.2) +
  scale_color_manual(values = c("Churned" = "red", "Not_Churned" = "blue")) +
  scale_linetype_manual(values = c("Churned" = "dashed", "Not_Churned" = "solid")) +
  labs(
    title = "Churned vs Retained Customers by Account Length",
    x = "Account Length (in months)",
    y = "Number of Customers"
  ) +
  theme_minimal() +
  theme(
    axis.title = element_text(face = "bold", size = 13),
    legend.title = element_blank(),
    legend.text = element_text(size = 12),
    plot.title = element_text(face = "bold", hjust = 0.5)
  )
```

# Churn Category Pie Chart

```{r}
# Step 1: Prepare data (only churned customers with category info)
churn_category_data <- df %>%
  filter(churn_label == "Yes", !is.na(churn_category)) %>%
  count(churn_category, name = "count") %>%
  mutate(
    percent = round(count / sum(count) * 100, 1),
    label = paste0(churn_category, "\n", percent, "%")
  )

# Step 2: Define accessible color palette
n_categories <- nrow(churn_category_data)
color_palette <- brewer.pal(n = n_categories, name = "Set2")

# Step 3: Pie chart
ggplot(churn_category_data, aes(x = "", y = count, fill = churn_category)) +
  geom_bar(stat = "identity", width = 1, color = "white") +
  coord_polar("y", start = pi/3) +
  geom_text(aes(label = paste0(percent, "%")), 
            position = position_stack(vjust = 0.5), 
            fontface = "bold", size = 4) +
  scale_fill_manual(values = color_palette) +
  theme_void() +
  labs(title = "Churn Categories") +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
    legend.title = element_blank(),
    legend.position = "right"
  )
```

# Churn Distribution

```{r}
# Convert to factor
df$churn_label <- factor(df$churn_label, levels = c("No", "Yes"))

# Plot churn distribution
ggplot(df, aes(x = churn_label)) +
  geom_bar(fill = "#D55E00") +
  labs(title = "Churn Distribution", x = "Churn Label", y = "Count")
```

# Churn Count by Contract Type

```{r}
ggplot(df, aes(x = contract_type, fill = churn_label)) +
  geom_bar(position = "dodge") +
  labs(title = "Churn Count by Contract Type", x = "Contract Type", y = "Count") +
  scale_fill_manual(values = c("#56B4E9", "#D55E00")) +
  theme(axis.text.x = element_text(angle = 15, hjust = 1))
```

# Monthly Charges by Churn Status (Boxplot)

```{r}
ggplot(df, aes(x = churn_label, y = monthly_charge, fill = churn_label)) +
  geom_boxplot() +
  labs(title = "Monthly Charges by Churn Status", x = "Churn Label", y = "Monthly Charge") +
  scale_fill_manual(values = c("#56B4E9", "#D55E00"))
```

# Churn by International Plan

```{r}
ggplot(df, aes(x = intl_plan, fill = churn_label)) +
  geom_bar(position = "dodge") +
  labs(title = "Churn by International Plan", x = "International Plan", y = "Count") +
  scale_fill_manual(values = c("#56B4E9", "#D55E00"))
```

# Avg Monthly GB Download by Contract Type

```{r}
df %>%
  group_by(contract_type) %>%
  summarise(avg_monthly_gb_download = mean(avg_monthly_gb_download, na.rm = TRUE)) %>%
  arrange(desc(avg_monthly_gb_download))
```

# View Sample Data

```{r}
head(df, 5)
```

# Drop Unused Columns

```{r}
# Drop columns not needed for modeling
df1 <- df[, !names(df) %in% c("customer_id", "churn_category", "churn_reason", "state", "phone_number")]

# Preview
head(df1, 5)
```

# Data Splitting and Label Encoding

```{r}
# Convert target to factor
df1$churn_label <- as.factor(df1$churn_label)

# Split into train/test sets with stratification
set.seed(123)
train_index <- createDataPartition(df1$churn_label, p = 0.8, list = FALSE)

train <- df1[train_index, ]
test <- df1[-train_index, ]

# Check class distribution
str(train$churn_label)
table(train$churn_label)
```

# Recheck Data Preview

```{r}
head(df1, 5)
```

# Preprocessing Recipe (One-Hot Encoding & Scaling)

```{r}
# Step 1: Define recipe
rec <- recipe(churn_label ~ ., data = train) %>%
  step_dummy(all_nominal_predictors(), one_hot = TRUE) %>%
  step_center(all_numeric_predictors()) %>%
  step_scale(all_numeric_predictors())

# Step 2: Fit recipe to training data
prep_rec <- prep(rec, training = train)

# Step 3: Transform both training and test sets
train_processed <- bake(prep_rec, new_data = train)
test_processed <- bake(prep_rec, new_data = test)

# Step 4: Glimpse processed data
glimpse(train_processed)
```

# 1. Cross-Validation Folds

```{r}
set.seed(123)
folds <- vfold_cv(train_processed, v = 5, strata = churn_label)
```

# 2. Model Specifications

```{r}
# Logistic Regression
log_model <- logistic_reg(penalty = tune(), mixture = tune()) %>%
  set_engine("glmnet") %>%
  set_mode("classification")

# SVM
svm_model <- svm_rbf(cost = tune(), rbf_sigma = tune()) %>%
  set_engine("kernlab") %>%
  set_mode("classification")

# Random Forest
rf_model <- rand_forest(mtry = tune(), trees = 500, min_n = tune()) %>%
  set_engine("ranger", importance = "impurity") %>%
  set_mode("classification")
```

# 3. Workflows

```{r}
# Define workflows
wf_log <- workflow() %>% add_model(log_model) %>% add_formula(churn_label ~ .)
wf_svm <- workflow() %>% add_model(svm_model) %>% add_formula(churn_label ~ .)
wf_rf  <- workflow() %>% add_model(rf_model)  %>% add_formula(churn_label ~ .)
```

# 4. Tuning Grids

```{r}
# Logistic Regression
log_grid <- grid_regular(penalty(), mixture(), levels = 5)

# SVM
svm_grid <- grid_regular(cost(), rbf_sigma(), levels = 5)

# Random Forest
mtry_range <- finalize(mtry(), train_processed %>% select(-churn_label))
rf_params <- parameters(mtry_range, min_n())
rf_grid <- grid_random(rf_params, size = 10)
```

> **Why ROC AUC over Accuracy?**  
> ROC AUC is more appropriate for imbalanced datasets. With ~27% churn rate, accuracy is misleading. A model that predicts “No churn” for all would still get ~73% accuracy, but zero utility in identifying churners. AUC evaluates ranking quality and is better for retention targeting.

# 5. Tune Models

```{r}
# Logistic Regression
log_tuned <- tune_grid(
  wf_log,
  resamples = folds,
  grid = log_grid,
  metrics = metric_set(roc_auc)
)

# SVM
svm_tuned <- tune_grid(
  wf_svm,
  resamples = folds,
  grid = svm_grid,
  metrics = metric_set(roc_auc)
)

# Random Forest
rf_tuned <- tune_grid(
  wf_rf,
  resamples = folds,
  grid = rf_grid,
  metrics = metric_set(roc_auc)
)
```

# 6. Compare & Select Best Models

```{r}
# Top results
show_best(log_tuned, metric = "roc_auc")
show_best(svm_tuned, metric = "roc_auc")
show_best(rf_tuned,  metric = "roc_auc")

# Select best configurations
log_best <- select_best(log_tuned, metric = "roc_auc")
svm_best <- select_best(svm_tuned, metric = "roc_auc")
rf_best  <- select_best(rf_tuned,  metric = "roc_auc")

# Print them
cat("Best Logistic Regression:\n"); print(log_best)
cat("Best SVM:\n"); print(svm_best)
cat("Best Random Forest:\n"); print(rf_best)
```

# 7. Final Fit and Evaluation on Test Set

```{r}
# Finalize best model (e.g., Random Forest)
final_rf <- finalize_workflow(wf_rf, rf_best)

# Split again for test evaluation (if needed)
final_split <- initial_split(train_processed, prop = 0.9, strata = churn_label)
final_fit <- last_fit(final_rf, split = final_split)

# Show final performance metrics
collect_metrics(final_fit)
```

# 8. ROC Curve

```{r}
# Get predictions
preds <- collect_predictions(final_fit)

# Compute ROC
roc_obj <- roc(response = preds$churn_label, predictor = preds$.pred_Yes, levels = rev(levels(preds$churn_label)))
roc_df <- data.frame(
  fpr = 1 - roc_obj$specificities,
  tpr = roc_obj$sensitivities
)

# Plot
ggplot(roc_df, aes(x = fpr, y = tpr)) +
  geom_line(color = "orange", size = 1.2) +
  geom_abline(linetype = "dashed", color = "darkblue") +
  labs(
    title = "Receiver Operating Characteristic (ROC)",
    x = "False Positive Rate",
    y = "True Positive Rate",
    caption = paste("AUC =", round(auc(roc_obj), 2))
  ) +
  theme_minimal(base_size = 14)
```

# 9. Model Evaluation Metrics

### Confusion Matrix

```{r}
conf_mat(preds, truth = churn_label, estimate = .pred_class)
```

### Accuracy

```{r}
accuracy(preds, truth = churn_label, estimate = .pred_class)
```

### Precision & Recall

```{r}
yardstick::precision(data = preds, truth = churn_label, estimate = .pred_class)
yardstick::recall(data = preds, truth = churn_label, estimate = .pred_class)
```

### All Metrics Summary

```{r}
yardstick::metrics(preds, truth = churn_label, estimate = .pred_class)
```

## Global Explanations

### 1. Feature Importance (VIP)

```{r}
# Extract final model from last_fit
final_model <- extract_fit_parsnip(final_fit)

# Visualize top 20 important features
vip(final_model, num_features = 20)
```

### 2. Partial Dependence Plot (PDP)

```{r}
# PDP for top feature (e.g., customer_service_calls)
pdp_plot <- partial(
  object = final_model,
  pred.var = "customer_service_calls",
  train = train_processed %>% select(-churn_label),
  prob = TRUE,
  which.class = 2  # class "Yes"
)

# Plot PDP
autoplot(pdp_plot) + ggtitle("Partial Dependence: customer_service_calls")
```

---

## Local Explanations

### 3. LIME: Explaining Individual Predictions

```{r}
# Re-prepare the recipe if not already done
rec_prep <- prep(rec, training = train)

# Build LIME explainer
explainer <- lime::lime(
  x = bake(rec_prep, new_data = train) %>% select(-churn_label),
  model = extract_fit_engine(final_model),
  bin_continuous = TRUE
)

# Define features to include in local explanation
custom_features <- c(
  "customer_service_calls",
  "contract_type_Month.to.Month",
  "monthly_charge",
  "intl_plan_yes",
  "total_charges"
)
```

#### LIME Explanation for Sample Observations

```{r}
# Loop through first 5 test observations
for (i in 1:5) {
  test_instance <- test[i:i, ]
  test_instance_processed <- bake(rec_prep, new_data = test_instance) %>% select(-churn_label)
  
  explanation <- lime::explain(
    x = test_instance_processed,
    explainer = explainer,
    labels = "Yes",
    n_features = length(custom_features),
    feature_select = "highest_weights",
    features = custom_features
  )
  
  # Plot explanation
  print(plot_features(explanation, ncol = 1))
}
```

---

## SHAP Explanations (Global + Local)

### 4. SHAP Values with `iml`

```{r}
# Wrap the final model with iml Predictor
predictor <- Predictor$new(
  model = final_model,
  data = train_processed %>% select(-churn_label),
  y = train_processed$churn_label,
  type = "prob"
)

# Compute SHAP values for the first test observation
shap <- Shapley$new(
  predictor,
  x.interest = test_processed %>%
    select(-churn_label) %>%
    slice(1)
)

# Select top features by mean absolute SHAP value
top_features <- shap$results %>%
  group_by(feature) %>%
  summarise(mean_phi = mean(abs(phi))) %>%
  arrange(desc(mean_phi)) %>%
  slice_head(n = 38) %>%
  pull(feature)

# Filter results to include only top features
shap$results <- shap$results %>%
  filter(feature %in% top_features)

# Visualize SHAP values for .pred_Yes
shap$results %>%
  filter(class == ".pred_Yes") %>%
  ggplot(aes(x = reorder(feature, phi), y = phi)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title = "SHAP Explanation for Churned Customers",
    x = "Feature",
    y = "SHAP Value"
  ) +
  theme_minimal()
```
