---
title: "Mitigating Class Imbalance with Tidymodels: Attrition Example"
format:
  html:
    self-contained: true
    toc: true
    toc-location: left
    toc-depth: 3
    code-fold: false
    code-tools: true
    theme: minty
    number-sections: false
execute:
  warning: false
  message: false
---

## Libraries

```{r}
library(tidymodels)
library(themis)
library(probably)
library(hardhat)
library(dplyr)
library(ggplot2)
library(modeldata)
```

## Load and Explore Data

```{r}
data(attrition)
attrition <- attrition %>%
  mutate(Attrition = factor(Attrition, levels = c("No", "Yes")))

glimpse(attrition)
table(attrition$Attrition)
attrition %>% select(-Attrition) %>% summary()
```

## Deal with Class Imbalance

### Preprocessing Methods

```{r}
# Downsampling
rec_down <- recipe(Attrition ~ ., data = attrition) %>%
  step_downsample(Attrition)

# Upsampling
rec_up <- recipe(Attrition ~ ., data = attrition) %>%
  step_upsample(Attrition)

# SMOTE
rec_smote <- recipe(Attrition ~ ., data = attrition) %>%
  step_smote(Attrition)
```

### Cost-sensitive Learning (Weights)

```{r}
class_counts <- table(attrition$Attrition)
attrition_weights <- attrition %>%
  mutate(.weight = 1 / class_counts[Attrition])
```

### Threshold Tuning

```{r}
threshold_metrics <- metric_set(roc_auc, pr_auc, accuracy, precision, recall, f_meas)
```

## Learning Flow

```{r}
set.seed(123)
split <- initial_split(attrition_weights, strata = Attrition)
train_data <- training(split)
test_data <- testing(split)

rec <- recipe(Attrition ~ ., data = train_data)

model_spec <- logistic_reg() %>%
  set_engine("glm")

workflow <- workflow() %>%
  add_model(model_spec) %>%
  add_recipe(rec)

# First, define the weight variable as a case weight in the recipe
rec <- recipe(Attrition ~ ., data = train_data) %>%
  update_role(.weight, new_role = "case_weight")

# Recreate the workflow using the updated recipe
workflow <- workflow() %>%
  add_model(model_spec) %>%
  add_recipe(rec)

# Now fit without passing weights explicitly — tidymodels will use the case_weight role
fit <- fit(workflow, data = train_data)

```

### Evaluate

```{r}
predictions <- predict(fit, test_data, type = "prob") %>%
  bind_cols(predict(fit, test_data)) %>%
  bind_cols(test_data)

metrics <- predictions %>%
  threshold_metrics(truth = Attrition, estimate = .pred_class, .pred_Yes)
metrics
```

## Tasks

- Best setup is the one maximizing F1, precision, and recall together.
- Compare metrics without balancing:
```{r}
rec_no_balance <- recipe(Attrition ~ ., data = train_data)
workflow_no_balance <- workflow() %>%
  add_model(model_spec) %>%
  add_recipe(rec_no_balance)
fit_no_balance <- fit(workflow_no_balance, data = train_data)

predictions_no_balance <- predict(fit_no_balance, test_data, type = "prob") %>%
  bind_cols(predict(fit_no_balance, test_data)) %>%
  bind_cols(test_data)

metrics_no_balance <- predictions_no_balance %>%
  threshold_metrics(truth = Attrition, estimate = .pred_class, .pred_Yes)
metrics_no_balance
```

- **Bonus**: Simulate a more severe imbalance

```{r}
attrition_severe <- attrition %>%
  filter(!(Attrition == "Yes" & runif(n()) < 0.5)) # drop half of the 'Yes'

table(attrition_severe$Attrition)
```

Repeat the above modeling steps with `attrition_severe` to see performance drop without balancing.
