
install.packages("kknn")
---
title: "KNN and the Curse of Dimensionality"
subtitle: "Regression"
format: html
code-fold: true
code-tools: true
embed-resources: true
page-layout: full
toc: true
toc-title: tasks
toc-location: left
toc-expand: 2
theme: minty
knitr:
  opts_chunk: 
    collapse: true
    R.options:
      message: false
      warning: false
    code-fold: false
---

```{r}
library(tidyverse)
library(tidymodels)
```

### Data Generation

This is your model:

$$
y_{i} = \beta_{1}x_{i1}+\epsilon_{i}
$$

where $\epsilon_{i} \sim N(0, \sigma^{2})$ and $x_{i1} \sim U(0,1)$. 

We pick $n = 1000$, $\beta_1 = 3$, and $\sigma^2 = 1$.

```{r}
set.seed(123)
n <- 1000
beta1 <- 3
sigma <- 1

x1 <- runif(n)
epsilon <- rnorm(n, mean = 0, sd = sigma)
y <- beta1 * x1 + epsilon

df <- tibble(x1 = x1, y = y)
```

---

### Basic Fit

Fit both linear regression and KNN regression to the data and evaluate their performance on the test set.

```{r}
set.seed(123)
split <- initial_split(df, prop = 0.8)
train <- training(split)
test <- testing(split)
```

#### Linear Model

```{r}
lm_model <- linear_reg() %>%
  set_engine("lm") %>%
  fit(y ~ x1, data = train)

lm_preds <- predict(lm_model, test) %>%
  bind_cols(test) %>%
  metrics(truth = y, estimate = .pred)

lm_preds
```

#### KNN Model with Tuning

```{r}
knn_spec <- nearest_neighbor(neighbors = tune()) %>%
  set_engine("kknn") %>%
  set_mode("regression")

knn_rec <- recipe(y ~ x1, data = train)

knn_wf <- workflow() %>%
  add_model(knn_spec) %>%
  add_recipe(knn_rec)

set.seed(123)
cv_folds <- vfold_cv(train, v = 5)

knn_grid <- grid_regular(neighbors(range = c(1, 30)), levels = 10)

knn_tune <- tune_grid(knn_wf, resamples = cv_folds, grid = knn_grid)

best_k <- select_best(knn_tune, metric = "rmse")

knn_final <- finalize_workflow(knn_wf, best_k)

knn_fit <- fit(knn_final, data = train)

knn_preds <- predict(knn_fit, test) %>%
  bind_cols(test) %>%
  metrics(truth = y, estimate = .pred)

knn_preds
```

---

### Add Dimensions

Now generate 30 random noise variables, progressively add them to the predictors, and re-fit both models.

```{r}
results <- tibble(p = integer(), model = character(), rmse = double())

for (p in 0:30) {
  if (p == 0) {
    X <- tibble(x1 = x1)
  } else {
    noise <- replicate(p, runif(n))
    X <- bind_cols(x1 = x1, as_tibble(noise))
  }
  df_p <- bind_cols(X, y = y)

  split <- initial_split(df_p, prop = 0.8)
  train <- training(split)
  test <- testing(split)

  # Linear model
  lm_fit <- linear_reg() %>%
    set_engine("lm") %>%
    fit(y ~ ., data = train)

  lm_rmse <- predict(lm_fit, test) %>%
    bind_cols(test) %>%
    metrics(truth = y, estimate = .pred) %>%
    filter(.metric == "rmse") %>%
    pull(.estimate)

  # KNN model
  rec <- recipe(y ~ ., data = train)
  wf <- workflow() %>%
    add_recipe(rec) %>%
    add_model(nearest_neighbor(neighbors = best_k$neighbors) %>%
                set_engine("kknn") %>%
                set_mode("regression"))

  knn_fit <- fit(wf, data = train)

  knn_rmse <- predict(knn_fit, test) %>%
    bind_cols(test) %>%
    metrics(truth = y, estimate = .pred) %>%
    filter(.metric == "rmse") %>%
    pull(.estimate)

  results <- results %>%
    add_row(p = p, model = "Linear", rmse = lm_rmse) %>%
    add_row(p = p, model = "KNN", rmse = knn_rmse)
}
```

---

### Is KNN Cursed?

Plot the performance of both models as the number of predictors increases.

```{r}
results %>%
  ggplot(aes(x = p, y = rmse, color = model)) +
  geom_line() +
  geom_point() +
  labs(title = "Model Performance vs Number of Predictors",
       x = "Number of Predictors",
       y = "RMSE",
       color = "Model") +
  theme_minimal()
```

---

### Conclusion

- As we increase the number of irrelevant features, **KNN** performance degrades rapidly.
- This is due to the **curse of dimensionality**: in high-dimensional space, distances between points become less meaningful.
- **Linear regression** is more robust to noise in this case, especially with uncorrelated predictors.
