---
title: "linear regression: back-to-back base vs tidy"
format: html
code-fold: true
code-tools: true
embed-resources: true
page-layout: full
toc: true
toc-title: quick tasks
toc-location: left
toc-expand: 2
theme: minty
knitr:
  opts_chunk: 
    collapse: true
    R.options:
    message: false
    warning: false
    code-fold: true
    echo: true
---

## initialization

```{r}
#| code-fold: false
library(tidyverse)
library(tidymodels)
library(ISLR2)
library(car)
```

### basic analysis

#### task 1

load the `Carseats` dataset from the `ISLR2` package, print the first few lines. Have a look at the help `?Carseats` to find out more about the variables.    

```{r}
# write your code here
```


#### task 2

Fit a regression model of `Sales` as function of `Price` and `Urban`. 
Use  `70%` of observation for training.

[base]{style="color:dodgerblue;"} 

```{r}
# write your code here
```

[tidy]{style="color:indianred;"}  

```{r}
# write your code here
```


#### task 3

Print the base output of the model.

- coefficient fit (how do you interpret $\hat{\beta}_{Price}$? and $\hat{\beta}_{UrbanYes}$?)

- goodness of fit

[base]{style="color:dodgerblue;"} 

```{r}
# write your code here
```

[tidy]{style="color:indianred;"}   

```{r}
# write your code here
```


#### task 4

This time use all the available predictors to predict  `Sales`. Use the same training set as before. 

[base]{style="color:dodgerblue;"} 

```{r}
# write your code here
```

[tidy]{style="color:indianred;"}   

```{r}
# write your code here
```

#### task 5

Remove the predictors with large `p-value`: plot the change in the  $R^2$? and adjusted $R^2$ as you remove the predictors with large p-value. (this is a home-made backward selection). Pick the model with the best $R^2_{adj}$.

[base]{style="color:dodgerblue;"} 

```{r}
# write your code here
```

[tidy]{style="color:indianred;"}   

```{r}
# write your code here
```

#### task 6

Compute the`rmse` test for both the full model and the reduced model. Which one is better?

[base]{style="color:dodgerblue;"} 

```{r}
# write your code here
```

[tidy]{style="color:indianred;"}   

```{r}
# write your code here
```

#### task 7

check the predictors for multicollinearity using the variance inflation factor (use `car::vif`)

[base]{style="color:dodgerblue;"} 

```{r}
# write your code here
```

[tidy]{style="color:indianred;"}   

```{r}
# write your code here
```

#### task 8

standardize the quantitative predictors. How do the coefficient change? (in `tydimodels`, use  `step_normalize()`)

[base]{style="color:dodgerblue;"} 

```{r}
# write your code here
```

[tidy]{style="color:indianred;"}   

```{r}
# write your code here
```



#### task 9

create a QQplot, and the residuals vs fitted values plot.

[base]{style="color:dodgerblue;"} 

```{r}
# write your code here
```

[tidy]{style="color:indianred;"}   

```{r}
# write your code here
```

## cross-validation and tuning

#### task 10

Create 5 folds off of the training set.

[base]{style="color:dodgerblue;"} 

```{r}
# write your code here
```

[tidy]{style="color:indianred;"}   

```{r}
# write your code here
```


#### task 11

fit a polynomial regression model of `Sales` as function of `Price`: tune-up the degree of the polynomial estimating the test `rmse`  via cross-validation.

[base]{style="color:dodgerblue;"} 

```{r}
# write your code here
```

[tidy]{style="color:indianred;"}   

```{r}
# write your code here
```

#### task 12

fit the polynomial regression model of `Sales` as function of `Price` using the best degree found in the previous task. Assess its performance on the test set.

[base]{style="color:dodgerblue;"} 

```{r}
# write your code here
```

[tidy]{style="color:indianred;"}   

```{r}
# write your code here
```


#### task 13

do the same using nearest neighbors regression. Tune the number of neighbors.

[base]{style="color:dodgerblue;"} 

```{r}
# write your code here
```

[tidy]{style="color:indianred;"}   

```{r}
# write your code here
```


#### task 14

select the model that provides the best performance on the test set.

```{r}
# write your code here
```

[tidy]{style="color:indianred;"}   

```{r}
# write your code here
```
