---
title: "Comparison of Hyperparameter Tuning Strategies"
format:
  html:
    self-contained: true
    toc: true
    toc-location: left
    toc-depth: 3
    code-fold: false
    code-tools: true
    theme: minty
    number-sections: false
execute:
  warning: false
  message: false
---

## Support vector machines on Spam Dataset

We use the **Spam dataset** for a binary classification task of identifying spam emails. The dataset is available in R via the `kernlab` package as `spam`.

### Preprocessing

We remove zero-variance predictors and normalize the features, which is critical for SVMs. The response variable is also ensured to be a factor.

```{r}
library(patchwork)
library(tidymodels)
library(kernlab)
library(finetune)
library(doParallel)
library(ggplot2)

data(spam)

spam <- spam %>% mutate(type = as.factor(type))

set.seed(123)
spam_split <- initial_split(spam, prop = 0.8, strata = type)
spam_train <- training(spam_split)
spam_test <- testing(spam_split)

spam_folds <- vfold_cv(spam_train, v = 3, strata = type)

spam_recipe <- recipe(type ~ ., data = spam_train) %>%
  step_zv(all_predictors()) %>%
  step_normalize(all_predictors())
```

### Model Specification

We use a radial basis function kernel SVM. The two main hyperparameters to tune are `cost` and `rbf_sigma`.

```{r}
svm_model <- svm_rbf(
  cost = tune(),
  rbf_sigma = tune()
) %>%
  set_engine("kernlab") %>%
  set_mode("classification")

svm_workflow <- workflow() %>%
  add_model(svm_model) %>%
  add_recipe(spam_recipe)
```

## Hyperparameter Tuning Strategies

We compare **Grid Search**, **Random Search**, and **Bayesian Optimization** using the same number of parameter combinations.

### Search Space

```{r}
grid_vals <- grid_regular(
  cost(range = c(-2, 2)), 
  rbf_sigma(range = c(-5, -1)), 
  levels = 3
)
n_combinations <- nrow(grid_vals)
```

### Grid Search

```{r}
set.seed(123)
grid_results <- tune_grid(
  svm_workflow,
  resamples = spam_folds,
  grid = grid_vals,
  metrics = metric_set(roc_auc),
  control = control_grid(save_pred = TRUE)
)
```

### Random Search

```{r}
set.seed(123)
random_results <- tune_grid(
  svm_workflow,
  resamples = spam_folds,
  grid = grid_random(cost(range = c(-1, 2)), rbf_sigma(range = c(-4, -1)), size = n_combinations),
  metrics = metric_set(roc_auc),
  control = control_grid(save_pred = TRUE)
)
```

### Bayesian Optimization

```{r}
param_info <- parameters(svm_model) %>%
  update(
    cost = cost(range = c(-1, 2)),
    rbf_sigma = rbf_sigma(range = c(-4, -1))
  )

set.seed(123)
bayes_results <- tune_bayes(
  svm_workflow,
  resamples = spam_folds,
  param_info = param_info,
  initial = 5,
  iter = n_combinations - 5,
  metrics = metric_set(roc_auc),
  control = control_bayes(save_pred = TRUE, verbose = TRUE)
)
```

## Results Visualization

We compare AUC performance in 2D: `cost` on x-axis, `rbf_sigma` on y-axis, AUC as color.

```{r}
plot_search <- function(results, title) {
  results %>%
    collect_metrics() %>%
    filter(.metric == "roc_auc") %>%
    ggplot(aes(x = cost, y = rbf_sigma, fill = mean)) +
    geom_point(size = 4, shape = 21, color = "black") +
    scale_x_log10() + scale_y_log10() +
    labs(title = title, fill = "AUC") +
    theme_minimal()
}

plot_search(grid_results, "Grid Search") +
plot_search(random_results, "Random Search") +
plot_search(bayes_results, "Bayesian Optimization")
```

### Conclusion

This analysis compares three common hyperparameter tuning strategies applied to a binary classification SVM. Visual inspection shows differences in the exploration behavior and highlights the effectiveness of Bayesian optimization.
