---
title: "Cars and classifications"
author: Alfonso
format: html
code-fold: true
code-tools: true
embed-resources: true
page-layout: full
toc: true
toc-title: tasks
toc-location: left
toc-expand: 2
theme: minty
knitr:
  opts_chunk: 
    collapse: true
    R.options:
    message: false
    warning: false
    code-fold: false
---

### Initialization

```{r}
#install.packages(c("tidyverse","tidymodels","discrim","MASS","naivebayes","kknn"))

library(tidyverse)
library(tidymodels)
library(discrim)
library(ISLR2)
library(kableExtra)
library(kknn)
tidymodels_prefer()

```

### pre-processing

-  From the Auto data set, transform the **mpg** variable into binary. Choose an appropriate threshold to define **mpg_hl** which indicates whether a car has high or low consumption (higher/lower than the chosen threshold).


- Standardize the variables.

### model fit and assessment

```{r}

```

- Identify the predictors that have the greatest discriminatory power using scatterplot or boxplot, such as the one below(yes, you do need to replicate the one below).

```{r}

```


- Creare un data set per il training dei modelli e uno per testarli.

- Create a data set for training the models and one for testing them.

- Perform a LDA, QDA, NaiveBayes and KNN  on the best predictors identified above: estimate the test error rate.

- Should you need to tune the hyperparameters, use cross-validation.
